CrypTool Script \\ Mathematics and Cryptography

### 2002-07-23 script/en/chapters/numbertheory.tex
be comments

Prime numbers can be factorised in a unique (``{\em trivial}'') way:
\begin{eqnarray}
5 & = & 1 * 5 \nonumber \\
17 & =  & 1 * 17 \nonumber \\
1,013 &  = & 1 * 1,013 \nonumber \\
1,296,409 & = & 1 * 1,296,409. \nonumber
\end{eqnarray}

@@

Examples of the decomposition of such numbers into prime factors:
\begin{eqnarray}
4 & = & 2*2  \nonumber \\
6 & = & 2*3  \nonumber \\
91 & = & 7*13  \nonumber \\
161 & = & 7*23  \nonumber \\
767 & = & 13*59  \nonumber \\
1,029 & = & 3 * 7^3  \nonumber \\
5,324 & = & 22 * 11^3.  \nonumber 
\end{eqnarray}

@@

If you allow numbers other than primes as factors, there are several ways of factorising integers and the {\em 
uniqueness} is lost:
$$60 = 1*60 = 2*30 = 4*15 = 5*12 = 6*10 = 2*3*10 = 2*5*6 = 3*4*5 = \cdots$$
The 1st fundamental theorem only appears to be obvious. We can construct numerous other sets of 
numbers\footnote{These sets are formed especially from the set of natural numbers. An example of this can be found in this \hyperlink{uniqueness}{script} on page 
\pageref{remFundTheoOfArithm} at the end of chapter 2.1.} for which numbers in the set {\em cannot} be expressed 
uniquely as a product of the prime numbers of the set.

@@

{\bf Example:}
\begin{eqnarray}
2 + 4 \equiv 6 \equiv 1 {\rm ~(mod~ } 5) ; \quad 1 + 1 \equiv 2 \equiv 2 {\rm ~(mod~ } 5),  \nonumber \\
2 * 4 \equiv 8 \equiv 3 {\rm ~(mod~ } 5) ; \quad 3 * 4 \equiv 12 \equiv 2 {\rm ~(mod~ } 5). \nonumber
\end{eqnarray}

@@

In modular arithmetic, raising to the power is defined as repeated multiplication -- as usual except that 
multiplication is now slightly different.  We can even apply the usual rules, such as:
\begin{eqnarray}
a^{b+c} & = & a^b * a^c,  \nonumber \\
(a^b)^c & = & a^{b*c} = a^{c*b} = (a^c)^b. \nonumber
\end{eqnarray}

@@

{\bf Example 1:}
$$ (4^3)^2 \equiv 64^2 \equiv 4096 \equiv 1 {\rm ~(mod~} 5). $$
\begin{quote}
(1) We can speed up\footnote{The time required to calculate the multiplication of two numbers normally 
depends on the length of the numbers.  We can observe this if we use the school method to calculate, for instance, 
$474*228$.  The time required increases quadratically, because we need to multiply $3*3$ numbers. The 
numbers become considerably smaller if we reduce the interim result.} the calculation by reducing the {\bf 
interim results} modulo $5$ but we need to take care because not everything then works in the same way as in 
standard arithmetic.
\begin{eqnarray}
(4^3)^2 & \equiv & (4^3{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & (64{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 4^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 16 \equiv 1 {\rm ~(mod~}5). \nonumber
\end{eqnarray}

@@

\subsubsection*{3.6.3.1 ~~~Fast calculation of high powers} \hypertarget{hohpot}{} \index{Power}
\addcontentsline{toc}{subsubsection}{~~~~~~~~~~3.6.3.1 ~Fast calculation of high powers}
RSA encryption and decryption\footnote{See \hyperlink{Kapitel_3_10}{chapter 3.10 Proof of the RSA procedure with Euler-Fermat} \index{RSA}
and \hyperlink{Kapitel_3_12}{chapter 3.12 The RSA procedure with actual numbers}.} entails calculating high powers modulo $m$. 
For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit long integer\index{Long integer} number range provided we calculate $a^n$ 
by actually multiplying a with itself $n$ times in line with the definition. In the case of extremely large numbers, even 
a fast computer chip would take longer than the age of the universe to calculate a single exponential. Luckily, there is 
an extremely effective shortcut for calculating exponentials (but not for calculating logarithms).

@@

and the squares (mod $103$) can be calculated beforehand
\begin{eqnarray*}
87^2 & \equiv & 50 {\rm ~(mod~}103),\\
87^4 & \equiv & 50^2 \equiv 28 {\rm ~(mod~}103), \\
87^8 & \equiv & 28^2 \equiv 63 {\rm ~(mod~}103), \\
87^{16} & \equiv & 63^2 \equiv 55 {\rm ~(mod~}103),\\
87^{32} & \equiv & 55^2 \equiv 38 {\rm ~(mod~}103).
\end{eqnarray*}
we have\footnote{%
  See \hyperlink{AppArith2}{Appendix D} for source code implementing the
  square and multiply method in Mathematica and Pari-GP, which can be used
  to reproduce the calculations above. }:
\label{SrcArith2}
\begin{eqnarray}
87^{43} & \equiv & 87^{32+8+2+1}{\rm ~(mod~}103) \nonumber \\
        & \equiv & 87^{32} * 87^8 * 87^2 * 87 {\rm ~(mod~}103) \nonumber \\ 
    & \equiv & 38 * 63 * 50 * 87 \equiv 85 {\rm ~(mod~}103). \nonumber
\end{eqnarray}

@@

If $J(m)$ cannot be determined\footnote{According to the first fundamental theorem of number 
theory\index{Number theory!fundamental theorem} and theorem \ref{thm-zth-phinum}, we determine $J(m)$ by reducing $m$ to prime factors.}, it is difficult to calculate 
the $c$-th root. This forms the basis for the security assumption used by the RSA encryption system (see 
Sub-section 4.3.1: \hyperlink{RSA procedure}{the RSA procedure} and \hyperlink{RSABeweis}{Sub-section 3.10 Proof of the RSA procedure}).
\end{itemize}
The time required for inverting addition and multiplication, on the other hand, is simply proportional to $\log m$ 
or $(\log m)^2$. Powers (for a number $x$ calculate $x^a$ with $a$ fixed) and exponents (for a number $x$ 
calculate $a^x$ with $a$ fixed) are therefore typical one way functions (See Overview of the
One way funnctions in this \hyperlink{OneWayFunktion1}{Script} and
\hyperlink{OneWayFunktion2}{artikel}). \index{One way function}

@@

Example for $a = 3$ modulo $10$ with $\mathbb{Z}_n^* = \{ 1, 3, 7, 9 \}$ :
\begin{eqnarray}
3 & \equiv & 3 * 1{\rm ~(mod~}10), \nonumber \\
9 & \equiv & 3 * 3{\rm ~(mod~}10), \nonumber \\
1 & \equiv & 3 * 7{\rm ~(mod~}10), \nonumber \\
7 & \equiv & 3 * 9{\rm ~(mod~}10). \nonumber 
\end{eqnarray}

@@

{\bf Example:}\par
Find the multiplicative inverse of $1579$ modulo $7351$.\\
According to Euler-Fermat:  $a^{J(n)} = 1 {\rm ~(mod~} n)$ for all $a$ in $\mathbb{Z}_n^*$.
If we divide both sides by $a$, we get: $a^{J(n) - 1} \equiv a^{-1} {\rm ~(mod~} n)$.
For the special case that the modulo is prime, we have $J(n) = p - 1$.
Therefore, the modular inverse is 
$$a^{-1} = a^{J(n) - 1} \equiv a^{(p-1)-1} \equiv a^{p-2} {\rm ~(mod~} p).$$
For our example, this means:
\begin{itemize}
\item[] Since the modulus $7351$ is prime, $p-2 = 7349$. \\
    $1579^{-1} \equiv 1579^{7349} {\rm ~(mod~} p).$
\end{itemize}
By cleverly breaking down the exponent, we can calculate this power relatively easily (see Section 3.6.3.1
\hyperlink{hohpot}{Fast calculation of high powers)}:
\begin{itemize}
\item[] $7349 = 4096 + 2048 + 1024 + 128 + 32 + 16 + 4 + 1$
\item[] $1579^{-1} \equiv 4716 {\rm ~(mod~} 7351).$
\end{itemize}

@@

In the first step we show that: $M^{e * d} \equiv M{\rm ~(mod~}p)$.
This results from the requirements and from Euler-Fermat (theorem \ref{thm-zth-fermateuler}).
Since $n=p*q$ and $J(p*q)=(p-1)*(q-1)$ and since $e$ and $d$ are selected in such a way that $e*d \equiv 1 {\rm ~(mod~}J(n))$, 
there is a whole number $k$ such that: $e*d = 1 + k*(p-1)*(q-1)$.
\begin{eqnarray}
M^{e * d}  & \equiv & M^{1+k*J(n)} \equiv M * M^{k*J(n)} \equiv M * M^{k*(p-1)*(q-1)}{\rm ~(mod~}p) \nonumber \\
           & \equiv & M * (M^{p-1})^{k*(q-1)}{\rm ~(mod~}p) \quad {\rm ~based~on~little~Fermat:~} 
                  M^{p-1} \equiv 1 {\rm~(mod~}p) \nonumber \\ 
           & \equiv & M * (1)^{k*(q-1)} {\rm~(mod~}p) \nonumber \\
       & \equiv & M {\rm ~(mod~}p) \nonumber
\end{eqnarray}
The requirement for using the simplified Euler-Fermat theorem (theorem \ref{thm-zth-fermat1}) was that $M$ and $p$ are relatively prime.

@@

\item However, to other relationships can be derived from the gcd:
For this, we need the set of equations for $a$ and $b$:
\begin{eqnarray}
 a & = & 1*a + 0*b \nonumber \\
 b & = & 0*a + 1*b, \nonumber
\end{eqnarray}
or, in matrix notation:
$$ \left(\begin{array}{c}a \\ b\end{array}\right) = 
   \left(\begin{array}{cc} 1 & 0 \\ 0 & 1 \end{array}\right) *
   \left(\begin{array}{c} a \\ b \end{array} \right).$$
We summarise this information in the extended matrix:
$$\left(\begin{array}{cccc} a & | & 1 & 0 \\ b & | & 0 & 1 \end{array} \right)$$
If we apply the above gcd algorithm to this matrix, we obtain the extended gcd algorithm:

@@

Since this algorithm only performs linear transformations, the same equations always apply
\begin{eqnarray}
 a & = & x_{1,1}*a + x_{1,2}*b \nonumber \\
 b & = & x_{2,1}*a + x_{2,2}*b, \nonumber
\end{eqnarray}
and we have the extended gcd equation at the end of the algorithm\footnote{By termination of the gcd algorithm,
the program variables $a$ and $b$ contain the values $a= 0$ and $b=gcd(a,b)$. Please keep in mind, that the program
variables are different to the numbers $a$ and $b$ and that they are only relevant for the scope of the algorithm.}:
$$gcd(a,b) = a*x_{2,1} + b*x_{2,2}.$$

@@

The property of closedness \index{Closedness} is always defined in relation to an operation in a set.
The following shows how to construct the ``closed set'' $G$ with respect to the operation $+ {\rm ~(mod~} 8)$ for a given initial set $G_0$:
\begin{eqnarray}
G_0 & = & \{ 2, 3 \} {\rm ~addition~of~the~numbers~in~} G_0 {\rm ~determines~further~numbers:} \nonumber \\
    & &    2 + 3 \equiv 5{\rm ~(mod~}8) = 5 \nonumber \\
    & &    2 + 2 \equiv 4{\rm ~(mod~}8) = 4 \nonumber \\
    & &    3 + 3 \equiv 6{\rm ~(mod~}8) = 6 \nonumber \\ 
G_1 & = & \{ 2, 3, 4, 5, 6 \} {\rm ~addition~of~the~numbers~in~} G_1 {\rm ~determines:}\nonumber \\
    & &    3 + 4 \equiv 7{\rm ~(mod~}8) = 7 \nonumber \\
    & &    3 + 5 \equiv 8{\rm ~(mod~}8) = 0 \nonumber \\
    & &    3 + 6 \equiv 9{\rm ~(mod~}8) = 1 \nonumber \\ 
G_2 & = & \{ 0, 1, 2, 3, 4, 5, 6, 7 \} {\rm ~~addition~of~the~numbers~in~} G_2 {~does~not~extend~the~set!} \nonumber \\
G_3 & = & G_2 {\rm ~~we~say:~} G_2 {\rm~is~closed~for~addition~~(mod~}8). \nonumber 
\end{eqnarray}
End of forming a closed set.
\vskip +12pt

>>>>

Prime numbers can be factorised in a unique (``{\em trivial}'') way:
\begin{eqnarray*}
5 & = & 1 * 5 \nonumber \\
17 & =  & 1 * 17 \nonumber \\
1,013 &  = & 1 * 1,013 \nonumber \\
1,296,409 & = & 1 * 1,296,409. \nonumber
\end{eqnarray*}

@@

Examples of the decomposition of such numbers into prime factors:
\begin{eqnarray*}
4 & = & 2*2  \nonumber \\
6 & = & 2*3  \nonumber \\
91 & = & 7*13  \nonumber \\
161 & = & 7*23  \nonumber \\
767 & = & 13*59  \nonumber \\
1,029 & = & 3 * 7^3  \nonumber \\
5,324 & = & 22 * 11^3.  \nonumber 
\end{eqnarray*}

@@

If you allow numbers other than primes as factors, there are several ways of factorising integers and the {\em 
uniqueness} is lost:
$$60 = 1*60 = 2*30 = 4*15 = 5*12 = 6*10 = 2*3*10 = 2*5*6 = 3*4*5 = \cdots$$
The 1st fundamental theorem only appears to be obvious. We can construct numerous other sets of 
numbers\footnote{These sets are formed especially from the set of natural numbers. An example of this can be found in this \hyperlink{uniqueness}{script} on page 
\pageref{remFundTheoOfArithm} at the end of chapter \ref{primesinmath}} for which numbers in the set {\em cannot} be expressed 
uniquely as a product of the prime numbers of the set.

@@

{\bf Example:}
\begin{eqnarray*}
2 + 4 \equiv 6 \equiv 1 {\rm ~(mod~ } 5) ; \quad 1 + 1 \equiv 2 \equiv 2 {\rm ~(mod~ } 5),  \nonumber \\
2 * 4 \equiv 8 \equiv 3 {\rm ~(mod~ } 5) ; \quad 3 * 4 \equiv 12 \equiv 2 {\rm ~(mod~ } 5). \nonumber
\end{eqnarray*}

@@

In modular arithmetic, raising to the power is defined as repeated multiplication -- as usual except that 
multiplication is now slightly different.  We can even apply the usual rules, such as:
\begin{eqnarray*}
a^{b+c} & = & a^b * a^c,  \nonumber \\
(a^b)^c & = & a^{b*c} = a^{c*b} = (a^c)^b. \nonumber
\end{eqnarray*}

@@

{\bf Example 1:}
$$ (4^3)^2 \equiv 64^2 \equiv 4096 \equiv 1 {\rm ~(mod~} 5). $$
\begin{quote}
(1) We can speed up\footnote{The time required to calculate the multiplication of two numbers normally 
depends on the length of the numbers.  We can observe this if we use the school method to calculate, for instance, 
$474*228$.  The time required increases quadratically, because we need to multiply $3*3$ numbers. The 
numbers become considerably smaller if we reduce the interim result.} the calculation by reducing the {\bf 
interim results} modulo $5$ but we need to take care because not everything then works in the same way as in 
standard arithmetic.
\begin{eqnarray*}
(4^3)^2 & \equiv & (4^3{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & (64{\rm ~(mod~}5))^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 4^2{\rm ~(mod~}5) \nonumber \\
            & \equiv & 16 \equiv 1 {\rm ~(mod~}5). \nonumber
\end{eqnarray*}

@@

\subsubsection{Fast calculation of high powers} \hypertarget{hohpot}{} \label{hohpot}\index{Power}
%\subsubsection*{3.6.3.1 ~~~Fast calculation of high powers} \hypertarget{hohpot}{} \index{Power}
%\addcontentsline{toc}{subsubsection}{~~~~~~~~~~3.6.3.1 ~Fast calculation of high powers}
RSA encryption and decryption\footnote{See \hyperlink{Kapitel_3_10}{chapter 3.10 Proof of the RSA procedure with Euler-Fermat} \index{RSA}
and \hyperlink{Kapitel_3_12}{chapter 3.12 The RSA procedure with actual numbers}.} entails calculating high powers modulo $m$. 
For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit long integer\index{Long integer} number range provided we calculate $a^n$ 
by actually multiplying a with itself $n$ times in line with the definition. In the case of extremely large numbers, even 
a fast computer chip would take longer than the age of the universe to calculate a single exponential. Luckily, there is 
an extremely effective shortcut for calculating exponentials (but not for calculating logarithms).

@@

and the squares (mod $103$) can be calculated beforehand
\begin{eqnarray*}
87^2 & \equiv & 50 {\rm ~(mod~}103),\\
87^4 & \equiv & 50^2 \equiv 28 {\rm ~(mod~}103), \\
87^8 & \equiv & 28^2 \equiv 63 {\rm ~(mod~}103), \\
87^{16} & \equiv & 63^2 \equiv 55 {\rm ~(mod~}103),\\
87^{32} & \equiv & 55^2 \equiv 38 {\rm ~(mod~}103).
\end{eqnarray*}
we have\footnote{%
  See \hyperlink{AppArith2}{Appendix D} for source code implementing the
  square and multiply method in Mathematica and Pari-GP, which can be used
  to reproduce the calculations above. }:
\label{SrcArith2}
\begin{eqnarray*}
87^{43} & \equiv & 87^{32+8+2+1}{\rm ~(mod~}103) \nonumber \\
        & \equiv & 87^{32} * 87^8 * 87^2 * 87 {\rm ~(mod~}103) \nonumber \\ 
    & \equiv & 38 * 63 * 50 * 87 \equiv 85 {\rm ~(mod~}103). \nonumber
\end{eqnarray*}

@@

If $J(m)$ cannot be determined\footnote{According to the first fundamental theorem of number 
theory\index{Number theory!fundamental theorem} and theorem \ref{thm-zth-phinum}, we determine $J(m)$ by reducing $m$ to prime factors.}, it is difficult to calculate 
the $c$-th root. This forms the basis for the security assumption used by the RSA encryption system (see 
Sub-section 4.3.1: \hyperlink{RSAVerfahren}{the RSA procedure} and \hyperlink{RSABeweis}{Sub-section 3.10 Proof of the RSA procedure}).
\end{itemize}
The time required for inverting addition and multiplication, on the other hand, is simply proportional to $\log m$ 
or $(\log m)^2$. Powers (for a number $x$ calculate $x^a$ with $a$ fixed) and exponents (for a number $x$ 
calculate $a^x$ with $a$ fixed) are therefore typical one way functions (See Overview of the
One way funnctions in this \hyperlink{OneWayFunktion1}{Script} and
\hyperlink{OneWayFunktion2}{artikel}). \index{One way function}

@@

Example for $a = 3$ modulo $10$ with $\mathbb{Z}_n^* = \{ 1, 3, 7, 9 \}$ :
\begin{eqnarray*}
3 & \equiv & 3 * 1{\rm ~(mod~}10), \nonumber \\
9 & \equiv & 3 * 3{\rm ~(mod~}10), \nonumber \\
1 & \equiv & 3 * 7{\rm ~(mod~}10), \nonumber \\
7 & \equiv & 3 * 9{\rm ~(mod~}10). \nonumber 
\end{eqnarray*}

@@

{\bf Example:}\par
Find the multiplicative inverse of $1579$ modulo $7351$.\\
According to Euler-Fermat:  $a^{J(n)} = 1 {\rm ~(mod~} n)$ for all $a$ in $\mathbb{Z}_n^*$.
If we divide both sides by $a$, we get: $a^{J(n) - 1} \equiv a^{-1} {\rm ~(mod~} n)$.
For the special case that the modulo is prime, we have $J(n) = p - 1$.
Therefore, the modular inverse is 
$$a^{-1} = a^{J(n) - 1} \equiv a^{(p-1)-1} \equiv a^{p-2} {\rm ~(mod~} p).$$
For our example, this means:
\begin{itemize}
\item[] Since the modulus $7351$ is prime, $p-2 = 7349$. \\
    $1579^{-1} \equiv 1579^{7349} {\rm ~(mod~} p).$
\end{itemize}
By cleverly breaking down the exponent, we can calculate this power relatively easily (see Section~\ref{hohpot}
\hyperlink{hohpot}{Fast calculation of high powers)}:
\begin{itemize}
\item[] $7349 = 4096 + 2048 + 1024 + 128 + 32 + 16 + 4 + 1$
\item[] $1579^{-1} \equiv 4716 {\rm ~(mod~} 7351).$
\end{itemize}

@@

In the first step we show that: $M^{e * d} \equiv M{\rm ~(mod~}p)$.
This results from the requirements and from Euler-Fermat (theorem \ref{thm-zth-fermateuler}).
Since $n=p*q$ and $J(p*q)=(p-1)*(q-1)$ and since $e$ and $d$ are selected in such a way that $e*d \equiv 1 {\rm ~(mod~}J(n))$, 
there is a whole number $k$ such that: $e*d = 1 + k*(p-1)*(q-1)$.
\begin{eqnarray*}
M^{e * d}  & \equiv & M^{1+k*J(n)} \equiv M * M^{k*J(n)} \equiv M * M^{k*(p-1)*(q-1)}{\rm ~(mod~}p) \nonumber \\
           & \equiv & M * (M^{p-1})^{k*(q-1)}{\rm ~(mod~}p) \quad {\rm ~based~on~little~Fermat:~} 
                  M^{p-1} \equiv 1 {\rm~(mod~}p) \nonumber \\ 
           & \equiv & M * (1)^{k*(q-1)} {\rm~(mod~}p) \nonumber \\
       & \equiv & M {\rm ~(mod~}p) \nonumber
\end{eqnarray*}
The requirement for using the simplified Euler-Fermat theorem (theorem \ref{thm-zth-fermat1}) was that $M$ and $p$ are relatively prime.

@@

\item However, to other relationships can be derived from the gcd:
For this, we need the set of equations for $a$ and $b$:
\begin{eqnarray*}
 a & = & 1*a + 0*b \nonumber \\
 b & = & 0*a + 1*b, \nonumber
\end{eqnarray*}
or, in matrix notation:
$$ \left(\begin{array}{c}a \\ b\end{array}\right) = 
   \left(\begin{array}{cc} 1 & 0 \\ 0 & 1 \end{array}\right) *
   \left(\begin{array}{c} a \\ b \end{array} \right).$$
We summarise this information in the extended matrix:
$$\left(\begin{array}{cccc} a & | & 1 & 0 \\ b & | & 0 & 1 \end{array} \right)$$
If we apply the above gcd algorithm to this matrix, we obtain the extended gcd algorithm:

@@

Since this algorithm only performs linear transformations, the same equations always apply
\begin{eqnarray*}
 a & = & x_{1,1}*a + x_{1,2}*b \nonumber \\
 b & = & x_{2,1}*a + x_{2,2}*b, \nonumber
\end{eqnarray*}
and we have the extended gcd equation at the end of the algorithm\footnote{By termination of the gcd algorithm,
the program variables $a$ and $b$ contain the values $a= 0$ and $b=gcd(a,b)$. Please keep in mind, that the program
variables are different to the numbers $a$ and $b$ and that they are only relevant for the scope of the algorithm.}:
$$gcd(a,b) = a*x_{2,1} + b*x_{2,2}.$$

@@

The property of closedness \index{Closedness} is always defined in relation to an operation in a set.
The following shows how to construct the ``closed set'' $G$ with respect to the operation $+ {\rm ~(mod~} 8)$ for a given initial set $G_0$:
\begin{eqnarray*}
G_0 & = & \{ 2, 3 \} {\rm ~addition~of~the~numbers~in~} G_0 {\rm ~determines~further~numbers:} \nonumber \\
    & &    2 + 3 \equiv 5{\rm ~(mod~}8) = 5 \nonumber \\
    & &    2 + 2 \equiv 4{\rm ~(mod~}8) = 4 \nonumber \\
    & &    3 + 3 \equiv 6{\rm ~(mod~}8) = 6 \nonumber \\ 
G_1 & = & \{ 2, 3, 4, 5, 6 \} {\rm ~addition~of~the~numbers~in~} G_1 {\rm ~determines:}\nonumber \\
    & &    3 + 4 \equiv 7{\rm ~(mod~}8) = 7 \nonumber \\
    & &    3 + 5 \equiv 8{\rm ~(mod~}8) = 0 \nonumber \\
    & &    3 + 6 \equiv 9{\rm ~(mod~}8) = 1 \nonumber \\ 
G_2 & = & \{ 0, 1, 2, 3, 4, 5, 6, 7 \} {\rm ~~addition~of~the~numbers~in~} G_2 {~does~not~extend~the~set!} \nonumber \\
G_3 & = & G_2 {\rm ~~we~say:~} G_2 {\rm~is~closed~for~addition~~(mod~}8). \nonumber 
\end{eqnarray*}
End of forming a closed set.
\vskip +12pt

### 2002-07-23 script/en/chapters/primes.tex
be comments

%\paragraph{What are prime numbers?}
\subsection*{What are prime numbers?}
\addcontentsline{toc}{subsection}{~~~~~~What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

@@

\subsection{Prime numbers in mathematics}

@@

Thus there is an {\em infinite} number of primes
\hyperlink{primhfk}{(Cross-reference: overview under 2.6.1 of the number of
prime numbers in various intervals)}.\par \vskip + 3pt

@@

The largest prime numbers known today have several hundred
thousand digits, which is too big for us to imagine. The number of
elementary particles in the universe is ``only'' estimated to be a
$80$-digit number \hyperlink{grosord}{(See: overview under 2.6.3
of various orders of magnitude / dimensions)}.

@@

Cross-reference:  \hyperlink{ntePrimzahl}{the table under 2.6.2} contains the precise values for the $n$th
prime numbers for selected $ n.$

@@

\hyperlink{primhfk}{(Cross-reference: overview under 2.6.1 of the number of
prime numbers in various intervals).}

@@

\newpage
\subsubsection{Number of prime numbers in various intervals}
\hypertarget{primhfk}{}

@@

\newpage
\subsubsection{Indexing prime numbers ($n$-th prime number)} \hypertarget{ntePrimzahl}{}

@@

\newpage
\subsubsection{Orders of magnitude / dimensions in reality}
In the description of cryptographic protocols and algorithms, numbers occur that
are so large or so small that they are inaccessible to our intuitive
understanding. It may therefore be useful to provide comparative numbers from
the real world that surrounds us so that we can develop a feeling for the
security of cryptographic algorithms. Some of the numbers listed below originate
from \cite{Schwenk1996} and \cite[p.18]{Schneier1996}.  \hypertarget{grosord}{}
\begin{tabbing}
Probability that you will be hijacked on your next flight:~~ \= abcdefhijk \=
\kill
Probability that you will be hijacked on your next flight \> $ 5.5 \cdot 10^{-6}
$\> \\
Probability of 6 correct numbers in the lottery \> $ 7.1 \cdot 10^{-8} $\> \\
Annual probability of being hit by lightning \> $ 10^{-7} $\> \\
Risk of being hit by a meterorite \> $ 1.6 \cdot 10^{-12} $ \> \\
--------------------------------------------------------------------------------
----------------------------------\\
Time until the next ice age (in years) \> $14000 $ \> $ (2^{14})$ \\
Time until the sun dies away (in years)   \> $10^{9} $ \> $(2^{30})$ \\
Age of the Earth (in years)\> $ 10^9 $ \> $(2^{30}) $  \\
Age of the universe (in years) \> $ 10^{10} $ \> $(2^{34}) $ \\
Number of the Earth's atoms            \> $10^{51} $ \> $ (2^{170}) $ \\
Number of the sun's atoms              \> $10^{57}$ \> $ (2^{190})$ \\
Number of atoms in the universe (without dark material)        \> $10^{77}$  \>
$ (2^{265})$ \\
Volume of the universe (in $cm^3$)     \> $10^{84}$ \> $(2^{280})$
\end{tabbing}

>>>>

\subsection{What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

@@

\subsection{Prime numbers in mathematics}\label{primesinmath}

@@

Thus there is an {\em infinite} number of primes
\hyperlink{primhfk}{(Cross-reference: overview under \ref{s:primhfk} of the number of
prime numbers in various intervals)}.\par \vskip + 3pt

@@

The largest prime numbers known today have several hundred
thousand digits, which is too big for us to imagine. The number of
elementary particles in the universe is ``only'' estimated to be a
$80$-digit number \hyperlink{grosord}{(See: overview under \ref{s:grosord}
of various orders of magnitude / dimensions)}.

@@

Cross-reference:  \hyperlink{ntePrimzahl}{the table under \ref{s:ntePrimzahl}} contains the precise values for the $n$th
prime numbers for selected $ n.$

@@

\hyperlink{primhfk}{(Cross-reference: overview under \ref{s:primhfk} of the number of
prime numbers in various intervals).}

@@

\newpage
\subsubsection{Number of prime numbers in various intervals}\label{s:primhfk}
\hypertarget{primhfk}{}

@@

\newpage
\subsubsection{Indexing prime numbers ($n$-th prime number)} \hypertarget{ntePrimzahl}{}\label{s:ntePrimzahl}

@@

\newpage
\subsubsection{Orders of magnitude / dimensions in reality}\label{s:grosord}
In the description of cryptographic protocols and algorithms, numbers occur that
are so large or so small that they are inaccessible to our intuitive
understanding. It may therefore be useful to provide comparative numbers from
the real world that surrounds us so that we can develop a feeling for the
security of cryptographic algorithms. Some of the numbers listed below originate
from \cite{Schwenk1996} and \cite[p.18]{Schneier1996}.  \hypertarget{grosord}{}
\begin{tabbing}
Probability that you will be hijacked on your next flight:~~ \= abcdefhijk \=
\kill
Probability that you will be hijacked on your next flight \> $ 5.5 \cdot 10^{-6}
$\> \\
Probability of 6 correct numbers in the lottery \> $ 7.1 \cdot 10^{-8} $\> \\
Annual probability of being hit by lightning \> $ 10^{-7} $\> \\
Risk of being hit by a meterorite \> $ 1.6 \cdot 10^{-12} $ \> \\
--------------------------------------------------------------------------------
----------------------------------\\
Time until the next ice age (in years) \> $14000 $ \> $ (2^{14})$ \\
Time until the sun dies away (in years)   \> $10^{9} $ \> $(2^{30})$ \\
Age of the Earth (in years)\> $ 10^9 $ \> $(2^{30}) $  \\
Age of the universe (in years) \> $ 10^{10} $ \> $(2^{34}) $ \\
Number of the Earth's atoms            \> $10^{51} $ \> $ (2^{170}) $ \\
Number of the sun's atoms              \> $10^{57}$ \> $ (2^{190})$ \\
Number of atoms in the universe (without dark material)        \> $10^{77}$  \>
$ (2^{265})$ \\
Volume of the universe (in $cm^3$)     \> $10^{84}$ \> $(2^{280})$
\end{tabbing}
